---
title: "Cleaning Asthma Data "
author: "Ava"
date: '(`r format(Sys.time(), "%B %d, %Y")`)'
output: 
  github_document: default
  html_notebook: default 
---

## Introduction
This notebook...

## Dependencies
This notebook requires...

```{r load-packages}
# tidyverse packages
library(dplyr)       # data wrangling
library(readr)       # read/write tabular data
library(stringr)     # work with strings
library(tidyr)       # data wrangling
# spatial packages
library(janitor)     # data wrangling
library(sf)          # spatial data tools
library(mapview)     # projections
library(tigris)      # census data wrangling 
# other packages
library(here)        # file path management
library(naniar)      # missing data
library(ggplot2)
library(viridis)      # palettes
library(leaflet)
library(tidycensus) 
library(sp)
library(ggplot2)   
library(RColorBrewer)  # color palettes
library(viridis)       # color palettes


```



## Load Data 2015
loading asthma data from MoPhim data base
asthma is for 2015

```{r load-data}
asthma <- read_csv(here("data", "asthma.csv"))
```


Getting rid of first three rows bc they are pointless text. 
```{r p1-q1}
asthma %>%
  slice(4:1029) -> asthma_sliced
```

Getting rid of third col with no data. aslo renaming cols to 'count' and 'zip'
```{r}
asthma_sliced %>%
  select(-X3) %>%
  rename(count = `Missouri EPHT Asthma`) %>%
  rename(zip = `Title:`)-> asthma_sliced
```

Changing X to NA, making the count variable numeric
```{r}
mutate(asthma_sliced, count = ifelse(count == "x", NA, count)) -> asthma_cleaned
```
```{r}
mutate(asthma_sliced, count = as.numeric(count)) -> asthma_cleaned
```
`

loading Mo zip codes 
```{r}
moZip <- zctas(state = "Missouri", year = 2010,
               cb = FALSE, class ="sf")
```

Changing object to SF
```{r}
st_as_sf(moZip) -> moZip
```

cleaning names, selecting just cols for zip and geometry, changing col name to zip
```{r}
moZip %>%
  clean_names() %>%
  select(zcta5ce10, geometry) %>%
  rename(zip = `zcta5ce10`)-> moZip_clean
```

## Load Data 2014
loading data for asthma 2014
```{r}
asthma_14<- read_csv(here("data", "asthma2014.csv"))
```

getting rid of first three rows bc they have nothing in them 
```{r}
asthma_14 %>%
  slice(4:1029) -> asthma14_sliced
```

getting rid of empty third col, renaming remaning cols
```{r}
asthma14_sliced %>%
  select(-X3) %>%
  rename(count = `Missouri EPHT Asthma`) %>%
  rename(zip = `Title:`) -> asthma14_sliced
```

changing X to NA, changing to numeric
```{r}
mutate(asthma14_sliced, count = ifelse(count == "x", NA, count)) -> asthma14_cleaned
```

```{r}
mutate(asthma14_sliced, count = as.numeric(count)) -> asthma14_cleaned
```
left join to asthma_cleaned by zip. Now table has data for both 2014 and 2015 
```{r}
left_join(asthma_cleaned, asthma14_cleaned, by = "zip") -> joined_14_15
```

## Load Data 2013
loading asthma data for 2013 
```{r}
asthma_2013 <- read_csv(here("data", "asthma2013.csv")) 
```
Getting rid of first 4 rows bc they have no data 
```{r}
asthma_2013 %>%
  slice(4:1029) -> asthma13_sliced
```

getting rid of third col, renaming other cols
```{r}
asthma13_sliced %>%
  select(-X3) %>%
  rename(count_13 = `Missouri EPHT Asthma`) %>%
  rename(zip = `Title:`) -> asthma13_sliced
```

changing X to NA, changing count to numeric 
```{r}
mutate(asthma13_sliced, count_13 = ifelse(count_13 == "x", NA, count_13)) -> asthma13_cleaned
```
```{r}
mutate(asthma13_sliced, count_13 = as.numeric(count_13)) -> asthma13_cleaned
```
left joing to joined_14_15, so table now has data for years 2013, 2014, and 2015 
```{r}
left_join(joined_14_15, asthma13_cleaned, by = "zip") -> joined_13_14_15
```

## Load Data 2012
loading data for 2012 
```{r}
asthma_2012 <- read_csv(here("data", "asthma2012.csv")) 
```

getting rid of empty cols
```{r}
asthma_2012 %>%
  slice(4:1029) -> asthma12_sliced
```

cleaning data
```{r}
asthma12_sliced %>%
  select(-X3) %>%
  rename(count_12 = `Missouri EPHT Asthma`) %>%
  rename(zip = `Title:`) -> asthma12_sliced
```

changing X to NA and changing count to numeric 
```{r}
mutate(asthma12_sliced, count_12 = ifelse(count_12 == "x", NA, count_12)) -> asthma12_cleaned
```
```{r}
mutate(asthma12_sliced, count_12 = as.numeric(count_12)) -> asthma12_cleaned
```
left joing to joined_13_14_15 by zip. Table now has data for 2012 through 2015 
```{r}
left_join(joined_13_14_15, asthma12_cleaned, by = "zip") -> joined12_15
```

## Load Data 2011
loading data for 2011
```{r}
asthma_2011 <- read_csv(here("data", "asthma2011.csv")) 
```
getting rid of empty cols
```{r}
asthma_2011 %>%
  slice(4:1029) -> asthma11_sliced
```

cleaning data
```{r}
asthma11_sliced %>%
  select(-X3) %>%
  rename(count_11 = `Missouri EPHT Asthma`) %>%
  rename(zip = `Title:`) -> asthma11_sliced
```

changing X to NA, and changing count to be numeric 
```{r}
mutate(asthma11_sliced, count_11 = ifelse(count_11 == "x", NA, count_11)) -> asthma11_cleaned
```

```{r}
mutate(asthma11_sliced, count_11 = as.numeric(count_11)) -> asthma11_cleaned
```

## Joining data for years 2011-2015
joining to joined_12_15 for years 2011-2015
```{r}
left_join(joined12_15, asthma11_cleaned, by = "zip") -> joined11_15
```

joining data with geometry by zip
```{r}
asthma_full_11_15 <- left_join(moZip_clean, joined11_15, by = "zip")
```
```{r}
asthma_full_11_15 %>%
rename(
  count_14 = count.y,
  count_15 = count.x
) -> asthma_full_11_15
```

## Zip codes for St Louis metro area

filtering for Zip codes in St Louis Metro Area, first filter gets range, Exclude gets rid of zip codes in that range that aren't included in the metro area. 
```{r}
asthma_full_11_15 %>%
  filter(zip >= (63005))%>%
  filter(zip <= (63390)) -> asthma_metro
```

  
```{r}
exclude <- c(63091, 63155, 63333, 63334, 63336, 63339, 63344, 63345, 63350, 63351, 63352, 63353, 63359, 63359, 63361, 63363, 63382, 63388, 63384, 63036, 63087, 63330)
```

```{r}
asthma_metro %>%
filter(as.character(zip) %in% exclude == FALSE) -> asthma_metro_2
```

changing NA's to 0s
```{r}
asthma_metro_2 %>%
  mutate(count_11 = ifelse(is.na(count_11) == TRUE, 0, count_11)) %>%
  mutate(count_12 = ifelse(is.na(count_12) == TRUE, 0, count_12)) %>%
  mutate(count_13 = ifelse(is.na(count_13) == TRUE, 0, count_13)) %>%
  mutate(count_14 = ifelse(is.na(count_14) == TRUE, 0, count_14)) %>%
  mutate(count_15 = ifelse(is.na(count_15) == TRUE, 0, count_15))-> asthma_metro_2
```


## Creating Count variable
Creating count for all 5 years

```{r}
asthma_metro_2 %>%
  group_by(zip) %>%
  mutate(total = sum(count_15, count_14, count_13, count_12, count_11)) %>%
  select(zip, total, geometry) -> asthma_count_total
```
rowwise() %>%
getting census data to get population for zip code tracts. using 5 year estimates from 2015
```{r}
acs <- load_variables(year = 2015, dataset = "acs5", cache = TRUE)
```

```{r}
pop <- get_acs(geography = "zip code tabulation area", year = 2015, variables = "B01003_001", survey = "acs5")
```

filtering for zip codes in st. louis metro area
```{r}
pop %>%
  filter(GEOID >= (63005))%>%
  filter(GEOID <= (63390)) ->pop
```
```{r}
pop %>%
  filter(as.character(GEOID) %in% exclude == FALSE)%>%
  rename(zip = GEOID,
         pop = estimate) %>%
  select(zip, pop) -> pop_metro
```

left join pop_metro to asthma_count_total, making variable for count 
```{r}
left_join(asthma_count_total, pop_metro, by = "zip") %>%
  mutate(count = (total/pop)*1000) %>%
 select(zip, count, geometry) -> count_by_pop
```

```{r}
mapview(count_by_pop)
```

creating ggplot map
```{r}
ggplot() +
    geom_sf(data = count_by_pop, mapping = aes(fill = count)) +
  scale_fill_distiller(palette = "Greens", trans ="reverse")
```

## saving csv file for Asthma data for years 2011-2015
```{r}
write.csv(count_by_pop, here("csv", "asthmaCount11_15.csv"))
```


## loading Data for % African American 

loading data for year 2015 (five year estimate)
```{r}
aa <- get_acs(geography = "zip code tabulation area", year = 2015, variables = "B02009_001", survey = "acs5")
```

Joining with pop data
```{r}
aa <- left_join(aa, pop, by = "GEOID") 
```

normalizing by pop
```{r}
aa %>%
mutate(percent = estimate.x/estimate.y) %>%
  rename(zip = GEOID) -> aa
```

join with asthma data 
```{r}
aa_metro <- left_join(count_by_pop, aa, by = "zip")
```

saving csv file for % african american 
```{r}
write.csv(aa_metro, here("csv", "percent_aa.csv"))
```

saving as shapefile
```{r}
dir.create(here("csv", "precent_aa"))
st_write(aa_metro, dsn = here("csv","precent_aa", "precent_aa.shp"), delete_dsn = TRUE)
```

## loading data for median income
loading five year estimate for 2015
```{r}
median_income <- get_acs(geography = "zip code tabulation area", year = 2015, variables = "B06011_001", survey = "acs5")
```

join with pop data 
```{r}
median_income <- left_join(median_income, pop, by = "GEOID") 

```

normalize by pop
```{r}
median_income%>%
mutate(percent = estimate.x/estimate.y) %>%
rename(zip = GEOID) -> median_income
```

join with asthma data 
```{r}
median_income <- left_join(count_by_pop, median_income, by = "zip")

```

saving csv file for median income 

```{r}
write.csv(median_income, here("csv", "median_income.csv"))
```

saving as shapefile
```{r}
dir.create(here("csv", "median_income"))
st_write(median_income, dsn = here("csv","median_income", "median_income.shp"), delete_dsn = TRUE)
```

## loading data for median home value 

loading 5 year estimate from 2015
```{r}
home_value <- get_acs(geography = "zip code tabulation area", year = 2015, variables = "B25077_001", survey = "acs5")
```

joining with pop data
```{r}
home_value <- left_join(home_value, pop, by = "GEOID") 
```

normalizing by pop
```{r}
home_value%>%
mutate(percent = estimate.x/estimate.y) %>%
rename(zip = GEOID) -> home_value
```

joining with asthma data
```{r}
home_metro <- left_join(count_by_pop, home_value, by = "zip")
```

Saving csv file for median home value

```{r}
write.csv(home_metro, here("csv", "median_home.csv"))
```

saving as shapefile 
```{r}
dir.create(here("csv", "home_value"))
st_write(home_metro, dsn = here("csv","home_value", "home_value.shp"), delete_dsn = TRUE)
```


## loading data for medicaide 
loading five year estimate from 2015
```{r}
medicaid <- get_acs(geography = "zip code tabulation area", year = 2015, variables = "B992707_001", survey = "acs5")

```

joining with pop
```{r}
medicaid <- left_join(medicaid, pop, by = "GEOID") 

```
normalizing by pop
```{r}
medicaid %>%
mutate(percent = estimate.x/estimate.y) %>%
rename(zip = GEOID) -> medicaid
```
joining with asthma data
```{r}
med_metro <- left_join(count_by_pop, medicaid, by = "zip")

```

saving csv file

```{r}
write.csv(med_metro, here("csv", "medicaid.csv"))
```

saving as shapefile
```{r}
dir.create(here("csv", "medicaid"))
st_write(med_metro, dsn = here("csv","medicaid", "medicaid.shp"), delete_dsn = TRUE)
```

## loading data for % under poverty line

loading five year estimates for 2015
```{r}
poverty <- get_acs(geography = "zip code tabulation area", year = 2015, variables = "B17001_002", survey = "acs5")
```
loading poverty total
```{r}
poverty_total <- get_acs(geography = "zip code tabulation area", year = 2015, variables = "B17001_001", survey = "acs5")
```

```{r}
poverty %>%
  select(GEOID, estimate) -> poverty
```


```{r}
poverty <- left_join(poverty, poverty_total, by = "GEOID") 
```
normalizing by povery total 
```{r}
poverty %>%
  mutate(density = estimate.x/estimate.y) %>%
  rename(zip = GEOID) -> poverty
```

joining with asthma data
```{r}
poverty_metro <- left_join(count_by_pop, poverty, by = "zip")
```


saving as csv file 
```{r}
write.csv(poverty_metro, here("csv", "poverty.csv"))

```

```{r}
dir.create(here("csv", "poverty"))
st_write(poverty_metro, dsn = here("csv","poverty", "poverty.shp"), delete_dsn = TRUE)
```


