---
title: "Ozone Research"
author: "Emily McCarthy"
date: '(`r format(Sys.time(), "%B %d, %Y")`)'
output: 
  github_document: default
  html_notebook: default 
---

## Introduction
This notebook will analyze how air quality changes over time in STL and SLC.

## Dependencies
This notebook requires...

```{r load-packages}
# tidyverse packages
library(dplyr)
library(stringr)
library(tibble)
library(readr)
library(tidyr)
library(leaflet)
library(mapview)

# other packages
library(here)
library(sf)
library(janitor)
library(naniar)

```

## Load Data
This notebook requires...

```{r load-data}
ozoneMo <- read_csv(here("data", "Ozone_2015.csv"))
ozone2012 <- read_csv(here("data", "Ozone_2012.csv"))
ozoneIL <- read_csv(here("data", "ILL_ozone_2015.csv"))
ozoneMad <- read_csv(here("data", "madison_ozone.csv"))
```

## Part 1

### Online Resources 


Sources used to find information about AQI and the ozone. 

0-300 SCALE

0-50 Safe 

50-100 Moderate for sensitive groups

100-150 Unhealthy for sensitive groups

151+ Unhealthy for all groups

https://ephtn.dhss.mo.gov/EPHTN_Data_Portal/airquality/pdf/aqibrochure.pdf
https://ephtn.dhss.mo.gov/EPHTN_Data_Portal/airquality/index.php
https://www3.epa.gov/aircompare/

### Data Cleaning

```{r}
ozoneMo %>%
  clean_names(case = "snake") -> ozoneMo

```

I will be subsetting the data in ozoneMO and ozone2012 so that only Metro West counties are included in these data. 

```{r}
ozoneMo %>%
  filter(county_code %in% c(510, 189, 99, 183)) %>%
  select(county_code, site_num, latitude, longitude, county_name, date_local, aqi, observation_count, observation_percent, x1st_max_value) -> ozoneMetro
```

I am now going to create geometry in the STL data set. 

```{r}
metro_sf <- st_as_sf(ozoneMetro, coords = c(x = "longitude", y = "latitude"), crs = 4269)
mapview(metro_sf)
```


EPA air quality open data: https://aqs.epa.gov/// https://www.epa.gov/outdoor-air-quality-data
Global open data index: https://index.okfn.org/

I will be mapping the location of the AQI station in Metro-East. 
 
```{r}
metro_sf <- select(metro_sf, "county_code", "county_name", "date_local", "observation_count", "observation_percent", "site_num", "aqi", "geometry")
mapview(metro_sf)
```

```{r}
# saving shapefile
st_write(metro_sf, dsn = here("data", "stlmetrolocations.shp"), delete_dsn = TRUE)
```

Subsetting all unhealthy (for sensitive groups) days throughout the year -- Days where AQI > 50

```{r}
metro_sf %>%
  filter(metro_sf$aqi >= 50) %>%
  group_by(site_num) %>%
  summarise(count = n()) -> unhealthy15
```

## Part 2 
Doing the exact same thing as above except for 2010-2014 

loading data
```{r}
o2014 <- read_csv(here("data", "O_2014.csv"))
o2013 <- read_csv(here("data", "O_2013.csv"))
o2012 <- read_csv(here("data", "O_2012.csv"))
o2011 <- read_csv(here("data", "O_2011.csv"))
```


```{r}
clean_asthma <- function(.data){
  
  .data <- janitor::clean_names(.data, case = "snake")
  .data <- dplyr::filter(.data, county_code %in% c(510, 189, 99, 183))
  .data <- dplyr::select(.data, date, aqs_site_id, daily_aqi_value, county_code, county, site_latitude, site_longitude)
  
  return(.data)
  
}
```


```{r}
o2014 <- clean_asthma(o2014)
o2013 <- clean_asthma(o2013)
o2012 <- clean_asthma(o2012)
o2011 <- clean_asthma(o2011)
```

Repeating the process for illinois -- reading data
```{r}
il2015 <- read_csv(here("data", "ILL_ozone_2015.csv"))
il2014 <- read_csv(here("data", "il2014.csv"))
il2013 <- read_csv(here("data", "il2013.csv"))
il2012 <- read_csv(here("data", "il2012.csv"))
il2011 <- read_csv(here("data", "il2011.csv"))
il2010 <- read_csv(here("data", "il2010.csv"))
```

```{r}
clean_illin <- function(.data){
  
  .data <- janitor::clean_names(.data, case = "snake")
  .data <- dplyr::filter(.data, county_code %in% c(119, 163))
  .data <- dplyr::select(.data, "date", "daily_aqi_value", "county_code", "county", "site_latitude", "site_longitude", "site_id")
  .data <- dplyr::rename(.data, aqs_site_id = site_id)
  
  return(.data)
  
}
```

selecting Madison and Saint Clair county
```{r}
il2015 <- clean_illin(il2015)
il2014 <- clean_illin(il2014)
il2013 <- clean_illin(il2013) 
il2012 <- clean_illin(il2012)
il2011 <- clean_illin(il2011)
il2010 <- clean_illin(il2010)
```

Joining illinois and missouri data
```{r}
metro_sf %>%
  rbind(metro_sf, il2015, deparse.level = 1) -> il2015

o2014 %>%
rbind(o2014, il2014, deparse.level = 1) -> il2014

o2013 %>%
rbind(o2013, il2013, deparse.level = 1) -> il2013

o2012 %>%
rbind(o2012, il2012, deparse.level = 1) -> il2012

o2011 %>%
rbind(o2011, il2011, deparse.level = 1) -> il2011
```

```{r}
st_as_sf(o2014, coords = c(x = "site_longitude", y = "site_latitude"), crs = 4269) -> o2014
st_as_sf(o2013, coords = c(x = "site_longitude", y = "site_latitude"), crs = 4269) -> o2013
st_as_sf(o2012, coords = c(x = "site_longitude", y = "site_latitude"), crs = 4269) -> o2012
st_as_sf(o2011, coords = c(x = "site_longitude", y = "site_latitude"), crs = 4269) -> o2011
```

creating count of unhealthy groups
```{r}
o2014 %>%
  filter(o2014$daily_aqi_value >= 50) %>%
  group_by(aqs_site_id) %>%
  summarise(count = n()) -> unhealthy14

o2013 %>%
  filter(o2013$daily_aqi_value >= 50) %>%
  group_by(aqs_site_id) %>%
  summarise(count = n()) -> unhealthy13

o2012 %>%
  filter(o2012$daily_aqi_value >= 50) %>%
  group_by(aqs_site_id) %>%
  summarise(count = n()) -> unhealthy12

o2011 %>%
  filter(o2011$daily_aqi_value >= 50) %>%
  group_by(aqs_site_id) %>%
  summarise(count = n()) -> unhealthy11

```

## Part 3: Kriging

```{r}
library(rspatial)
```

```{r}
unhealthy15 <- as_Spatial(unhealthy15)
```

```{r}
library(gstat)
gs <- gstat(formula=count~1, locations=unhealthy15)
v <- variogram(gs, width=20)
head(v)
plot(v)
```

Getting stuck on code below

```{r}
fve <- fit.variogram(v, vgm(85, "Exp", 75, 20))
fve
plot(variogramLine(fve, 400), type='l', ylim=c(0,120))
points(v[,2:3], pch=20, col='red')
```


#### Pre-SP CONVERSION

```{r}
as_Spatial(unhealthy15)
```

```{r}
mutate(unhealthy15, count = as.numeric(count), site_num = as.numeric(site_num)) -> unhealthy15n
```


```{r}
library(rspatial)
unhealthy15n$prec <- plot(sort(unhealthy15n$prec), ylab='Unhealthy Days', las=1, xlab='site_num')
```

```{r}
library(rspatial)
x <- sp_data("airqual")
x$OZDLYAV <- x$OZDLYAV * 1000
```


```{r}
library(gstat)
gs <- gstat(formula=OZDLYAV~1, locations=unhealthy15n)
v <- variogram(gs, width=20)
head(v)
plot(v)
```


```{r}
library(sp)
library(gstat)
gs <- gstat(formula=prec~1, locations=unhealthy15, nmax=5, set=list(idp = 0))
nn <- interpolate(r, gs)
## [inverse distance weighted interpolation]
nnmsk <- mask(nn, vr
plot(nnmsk)
```

```{r}
library(gstat)
```


```{r}
#Variogram from R pdf

lzn.vgm = variogram(log(count)~1, unhealthy15)
lzn.vgm
```

```{r}
lzn.fit = fit.variogram(lzn.vgm, model = vgm(1, "Sph", 900, 1))
lzn.fit
```

```{r}
plot(lzn.vgm, lzn.fit)
```

```{r}
#lznr.vgm = variogram(log(count)~sqrt(dist), unhealthy15n)
data(unhealthy15n.grid)
```

```{r}
library(stars)
lzn.kriged = krige(log(count)~1, unhealthy15n, unhealthy15n, model = lzn.fit)
```
```{r}
library(sp)
spplot(lzn.kriged["var1.pred"])
```
```{r}
lzn.idw = idw(count~1, unhealthy15n, unhealthy15n)
spplot(lzn.idw["var1.pred"], main = "u15 idw")
```

## Plotting

```{r}
ggplot() +
    geom_sf(data = unhealthy14, mapping = aes(fill = count)) +
  scale_fill_distiller(palette = "Greens", trans ="reverse")
```

