---
title: "Redlining Comparisons"
author: "Emily McCarthy"
date: '(`r format(Sys.time(), "%B %d, %Y")`)'
output: 
  github_document: default
  html_notebook: default 
---

## Introduction
This notebook will
- replace Ava's athsma data comparisions with redlining comparisons
- summarize 50 words the relationship between redlining and comtemporary demographic %black/%poverty

## Dependencies
This notebook requires...

```{r load-packages}
# tidyverse packages
library(dplyr)       # data wrangling
library(readr)       # read/write tabular data
library(stringr)     # work with strings
library(tidyr)       # data wrangling
# spatial packages
library(janitor)     # data wrangling
library(sf)          # spatial data tools
library(mapview)     # projections
library(tigris)      # census data wrangling 
# other packages
library(here)        # file path management
library(naniar)      # missing data
library(ggplot2)
library(viridis)      # palettes
library(leaflet)
library(tidycensus) 
library(sp)
library(ggplot2)   
library(RColorBrewer)  # color palettes
library(viridis)       # color palettes

```

## Load Data
This notebook requires...

```{r load-data}
redlining <- st_read(here("data", "clean", "STL_HOLC_zip40.geoJSON"))
```

Removing all zip codes that were not graded.

```{r p1-q1}
redlining <- filter(redlining, is.na(total_graded)== FALSE)
```

Loading population data

```{r}
acs <- load_variables(year = 2015, dataset = "acs5", cache = TRUE)
pop <- get_acs(geography = "zip code tabulation area", year = 2015, variables = "B01003_001", survey = "acs5")
```

Subsetting for the STL Metro Area

```{r}
pop %>%
  filter(GEOID >= (63005))%>%
  filter(GEOID <= (63390)) -> pop
```

```{r}
pop %>%
  rename(ZCTA = GEOID,
         pop = estimate) %>%
  select(ZCTA, pop) -> pop_metro
```

I am going to reduce columns in the redlining set to only show c and d grades- 

```{r}
redlining%>%
  select(ZCTA, prop_cd)  -> redlining_cd
```

renaming col ZCTA to zip

```{r}
redlining_cd %>%
  rename(zip = ZCTA) -> redlining_cd
```

joining C and D grades with population data

```{r}
cd <- left_join(redlining_cd, pop_metro, by = "ZCTA")
```

normalizing by population
```{r}
cd %>%
mutate(percent = prop_cd/pop) %>%
  rename(zip = ZCTA) -> cd
```

saving csv for redlining grades C and D
```{r}
write.csv(cd, here("redlining", "percent_cd.csv"))
```

saving as shapefile 
```{r}
dir.create(here("redlining", "percent_cd"))
st_write(cd, dsn = here("redlining", "percent_cd", "percent_cd.shp"), delete_dsn = TRUE)
```

#Up next: 
-import percent_aa.shp
-redlining and athma (1 plot)
-redlining and race (1 plot)

Importing .shp for comparison
```{r}
shapefile_aa <- st_read(here("csv", "precent_aa", "precent_aa.shp"), stringsAsFactors = FALSE)
```

Joining redlining info to athasma/AA data

```{r}
csv_aa <- read_csv(here("csv", "percent_aa.csv"))
```
```{r}
st_geometry(redlining_cd) <- NULL
```
```{r}
redlining_comparison <- left_join(redlining_cd, shapefile_aa, by = "zip")
```

Now that my data are joined I will make comparisons using ggplot

```{r}
ggplot(redlining_comparison, aes(x = prop_cd, y = count))+ geom_point(shape = 21)+
            labs(x = "Percent of redling grade C or D", y = "asthma") +
   geom_smooth(method='lm')
```

```{r}
ggplot(redlining_comparison, aes(x = prop_cd, y = percent))+ geom_point(shape = 21)+
            labs(x = "Percent of redlining grade C or D", y = "Percent African American") +
   geom_smooth(method='lm')
```

